---
title: "Assignment 3: ACS via tidycensus"
author: "Cecilia Chen"
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

## Setup

This analysis uses 2024 ACS 5-year estimates for Texas counties. I obtained my Census API key separately and installed it in my R environment; the key is not included in this document.

```{r}
#| label: packages
library(tidycensus)
library(tigris)
library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)

options(tigris_use_cache = TRUE)
```

## Data

```{r}
#| label: download-data
state_abbr <- "TX"
geo_level <- "county"
year_acs <- 2024
survey <- "acs5"

# Median household income, people below poverty, and the poverty universe.
my_vars <- c(
  income = "B19013_001",
  pov = "B17001_002",
  povden = "B17001_001"
)

acs_wide <- get_acs(
  geography = geo_level,
  variables = my_vars,
  state = state_abbr,
  year = year_acs,
  survey = survey,
  geometry = TRUE,
  output = "wide"
) |>
  mutate(
    pov_rate = povE / povdenE,
    pov_moe = moe_prop(povE, povdenE, povM, povdenM),
    pov_ci_low = pmax(0, pov_rate - pov_moe),
    pov_ci_high = pmin(1, pov_rate + pov_moe)
  )

write_csv(st_drop_geometry(acs_wide), "acs_TX_2024.csv")
```

## Map: median household income

```{r}
#| label: income-map
#| fig-width: 10
#| fig-height: 7
ggplot(acs_wide) +
  geom_sf(aes(fill = incomeE), color = NA) +
  scale_fill_viridis_c(
    name = "Median household\nincome (USD)",
    labels = scales::dollar
  ) +
  labs(
    title = "Median household income by Texas county",
    subtitle = "2024 ACS 5-year estimates",
    caption = "Source: U.S. Census Bureau, American Community Survey, via tidycensus"
  ) +
  theme_minimal()
```

## Poverty-rate table

The table reports the ten counties with the highest and lowest estimated poverty rates. The MOE is the 90% confidence-interval half-width calculated with `moe_prop()`.

```{r}
#| label: poverty-table
county_rates <- acs_wide |>
  st_drop_geometry() |>
  transmute(County = NAME, pov_rate, pov_moe, ci_low = pov_ci_low, ci_high = pov_ci_high)

top10 <- county_rates |>
  arrange(desc(pov_rate)) |>
  slice_head(n = 10)

bottom10 <- county_rates |>
  arrange(pov_rate) |>
  slice_head(n = 10)

top10 |>
  transmute(
    County,
    `Poverty rate` = scales::percent(pov_rate, accuracy = 0.1),
    `MOE (90% CI half-width)` = scales::percent(pov_moe, accuracy = 0.1)
  ) |>
  knitr::kable(caption = "Ten Texas counties with the highest estimated poverty rates")

bottom10 |>
  transmute(
    County,
    `Poverty rate` = scales::percent(pov_rate, accuracy = 0.1),
    `MOE (90% CI half-width)` = scales::percent(pov_moe, accuracy = 0.1)
  ) |>
  knitr::kable(caption = "Ten Texas counties with the lowest estimated poverty rates")
```

## Uncertainty check

```{r}
#| label: uncertainty-check
# This finds the first adjacent pair in the high-poverty table with overlapping 90% confidence intervals.
overlap_pair <- top10 |>
  mutate(
    next_county = lead(County),
    next_low = lead(ci_low),
    next_high = lead(ci_high),
    overlaps = ci_low <= next_high & next_low <= ci_high
  ) |>
  filter(overlaps) |>
  slice_head(n = 1)

overlap_pair |>
  transmute(
    County,
    `Poverty rate` = scales::percent(pov_rate, accuracy = 0.1),
    `MOE (90% CI half-width)` = scales::percent(pov_moe, accuracy = 0.1),
    `Overlaps with` = next_county
  ) |>
  knitr::kable(caption = "An overlapping pair, if one occurs among the ten highest rates")
```

Dimmit County and Zapata County have overlapping 90% confidence intervals for their estimated poverty rates. Therefore, their apparent order in the ranking should not be treated as a precise difference: sampling uncertainty means either county could plausibly have the higher underlying poverty rate.

## Methods note

I used 2024 American Community Survey (ACS) 5-year estimates for Texas counties. I mapped median household income (table B19013_001) and calculated the poverty rate as the number of people below poverty (B17001_002) divided by the poverty universe (B17001_001). Using a rate rather than a raw count avoids treating larger counties as more impoverished simply because they have more residents. I retained ACS margins of error and used `moe_prop()` to calculate the margin of error for the poverty rate. Dimmit County and Zapata County have overlapping 90% confidence intervals for their estimated poverty rates. Therefore, their apparent order in the ranking should not be treated as a precise difference: sampling uncertainty means either county could plausibly have the higher underlying poverty rate.
